---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, fig.path = "figures/")
options(scipen = 1, digits = 2)
```

## Loading and preprocessing the data

```{r loaddata}
data <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric"))
```


## What is mean total number of steps taken per day?

```{r meanTotalNumberOfSteps}
library(plyr)

# calculate the total steps per date
sum_steps <- ddply(data, c("date"), summarize,total_steps=sum(steps))
hist(sum_steps$total_steps,
     xlab = "Number of Steps",
     main = "Histogram of the Total No. of Steps Taken Each Day",
     col="yellow")
```

                Figure 1: Histogram of the Total No. of Steps Taken Each Day

```{r meanCompute}
mean_ <- mean(sum_steps$total_steps, na.rm = T)
median_ <- median(sum_steps$total_steps, na.rm = T)
```


The mean total number of steps taken per day is `r mean_`.

The median total number of steps taken per day is `r median_`.


## What is the average daily activity pattern?

```{r aveDailyPattern}
library(plyr)

# calculate the mean steps per interval across all dates
mean_data <- ddply(data,
                   c("interval"),
                   summarize,
                   mean_steps=mean(steps, na.rm = T))

plot(mean_data$interval,
     main = "Time Series Plot of the 5-Minute Interval\nand the Average No. of Steps Taken\n(Across All Days)",
     mean_data$mean_steps,
     ylab = "Ave. No. of Steps Taken", xlab = "5-Minute Interval", type="l")
```

                Figure 2: Time Series Plot of the 5-Minute Interval and the Average No. of Steps Taken

```{r getMaxInterval}
# getting 5-Min interval with the maximum average number of steps
max_row <- mean_data[mean_data$mean_steps == max(mean_data$mean_steps), ]
max_5 <- max_row[, c("interval")]
max_5_val <- max_row[, c("mean_steps")]
```


`r max_5` is the 5-min interval that contains the maximum number of steps, with the value
of `r max_5_val` (on average across all the days).


## Imputing missing values

```{r countNArows}

# original row count - row count with missing data omitted
na_count <- nrow(data) - nrow(na.omit(data))
```

The total number of missing values in the dataset is `r na_count` (number of rows with NAs).

```{r imputeMissingValues}
library(plyr)

# Strategy for filling in of missing values in dataset
new_data <- transform(data, 
                      steps = ifelse(is.na(steps), 
                                     floor(ave(steps,
                                               interval,
                                     FUN = function(x) mean(x, na.rm = TRUE))), 
                                     steps))

new_na_count <- nrow(data) - nrow(na.omit(new_data))

new_sum_steps <- ddply(new_data,
                       c("date"),
                       summarize,
                       total_steps=sum(steps, na.rm = F))

hist(new_sum_steps$total_steps,
     xlab = "Number of Steps",
     main = "Histogram of the Total No. of Steps Taken Each Day\n(Imputed Missing Data)",
     col="red")
```

                Figure 3: Histogram of the Total No. of Steps Taken Each Day (Imputed Missing Data)

```{r imputeComputeNew}
new_mean <- mean(new_sum_steps$total_steps, na.rm = T)
new_median <- median(new_sum_steps$total_steps, na.rm = T)
```


The new mean total number of steps taken per day is `r new_mean`.
The new median total number of steps taken per day is `r new_median`.
Both of these are different from the first computed values with still missing data.
The imputation affected the estimates of the total daily number of steps,
thus adding more amount of information available for the analysis.


## Are there differences in activity patterns between weekdays and weekends?

#### "Time Series Plot of the 5-Minute Interval and the Average No. of Steps Taken (Across All Weekend or Weekday Days)"

```{r weekEndDayActivity}
library(plyr)
library(ggplot2)

# New factor variable: whether a data is weekday or weekend
new_data$weekDayEnd <- ifelse(!weekdays(as.Date(new_data$date)) %in% c("Saturday", "Sunday"),
                              "weekday", "weekend")

n_data <- ddply(new_data, c("interval","weekDayEnd"), summarize, mean_steps=mean(steps))

print(ggplot(n_data, aes(interval, mean_steps)) + 
      geom_line() + 
      facet_grid(weekDayEnd~.) + 
      xlab("Interval") +
      ylab("Number of Steps"))

```

                      Figure 4: Time Series Plot of the 5-Minute Interval and the Average No. of Steps Taken


Figure 4 above shows that there is a significant difference in the activity patterns between weekdays and weekends. 
